Constructing Biological Regulatory Networks from Process Hitting models

### 2013-08-03 parts-asp.tex
all source files in the same folder (TCS submission)

\begin{example}\label{ex:asp-atom}
Consider a PH model such as \pref{ex:infer-param-runningPH-1}:
in order to state the existence of each component, we use a predicate called $\component$ with two arguments:
\begin{align*}
  &\component(x, n).
\end{align*}
where $x$ is the name of the component and $l_x = n$ is its maximum level of expression.
\end{example}

An ASP program is a set of rules as described above.
Solving an ASP program means finding an \emph{answer set}, which is a minimal set of atoms that respect all the rules.
In order to formally define this notion of answer set,
let us define a \emph{definite rule} as a rule with no negation of atoms (noted “$\nota$” above),
and a \emph{definite program} as a program containing only definite rules.
Let $S$ be a set of atoms: a definite rule is satisfied by $S$ if $\atom{H} \in S$,
or if $\exists i \in \segm{1}{n}, A_i \notin S$.
Given this definition of satisfaction, we define the answer set of a definite program $\Pi$
as the (unique) minimal set $S$ of atoms that satisfies all the rules in $\Pi$.

In order to consider the general case,
for all non-definite program $\Pi$ and set $S$ of atoms, we denote $\Pi^S$ the reduct of $\Pi$ w.r.t.~$S$, defined from $\Pi$ by:
\begin{enumerate}
  \item deleting all the rules that have a negation of atom $\nota \atom{B}_i$ in the body where $\atom{B}_i \in S$, and
  \item removing all negations of atoms in the bodies of the remaining rules.
\end{enumerate}
Then, a set of atoms $S$ is an answer set of a non-definite program $\Pi$ if it is the answer set of $\Pi^S$, which is a definite program.
We note that several answer sets can be solution to the same non-definite program, and in practice the solver can be directed to enumerate them all.

@@

\subsection{Facts}
A rule with no $body$ part is called a \emph{fact}, and its $head$ atom consequently has to belong to all answer sets.
For instance, the information describing the studied model (the original PH model and the inferred IG and parameters) are expressed in ASP using facts.
In particular, the predicate $\component$ allows to define all components belonging to the inferred IG.
Thus, an atom $\component(a, m)$ states that $a$ is a component of the IG and that $l_a = m$.

@@

\subsection{Variables}
To describe the sets of all expression levels of each component (\ie the set $\segm{0}{l_a}$ for each $a \in \Gamma$),
one can use atoms of the form $\componentlevels(a, k)$ to state that $k \in \segm{0}{l_a}$.
Variables here come in handy to enumerate each possible constant $k$ for each component $a$:
before solving, any rule containing variables is grounded, that is replaced by an equivalent set of rules with constants only.
%during the solving, any rule containing variables is duplicated in order to replace each variable by all the possible constants it could represent.
The following rule, for example, contains three variables ($\var{A}$, $\var{K}$ and $\var{M}$) and enumerates the set of possible expression levels of each component in the system:
\begin{align*}
  \componentlevels(A, K) \la \component(\var{A}, \var{M}), 0 \leq K \leq M.
\end{align*}
where the notation “$\leq$” stands for a shortcut in ASP which has the same meaning as the mathematical operator.

@@

\subsection{Constraints}\label{sssec:constraints}
Finally, a constraint is a rule with no $head$ part:
\begin{align*}
  \la \atom{A}_1, \atom{A}_2, \dots, \atom{A}_n, \nota \atom{B}_1, \nota \atom{B}_2, \dots, \nota \atom{B}_m.
\end{align*}
A constraint is satisfied only if its $body$ is not satisfied,
which thus allows to invalidate answer sets containing some unwanted combinations of atoms.
In the scope of parameters enumeration, for example, constraints are especially useful to filter parametrizations
that contain non-interval parameters, or that do not respect the assumptions of \pref{ssec:admissible-K}.
Indeed, suppose that:
\begin{itemize}
  \item $\lessactive(a, p, q)$ states that $\omega_p$ is a set of resources of $a$ with (loosely) less activators and more inhibitors than $\omega_q$;
  \item $\paraminf(a, p, q)$ states that: $K_{a,\omega_p} \leqsegm K_{a,\omega_q}$.
\end{itemize}
Then, the monotonicity assumption (\pref{pro:param_enum_monotonicity}) is formulated as the following constraint:
\begin{align*}
  \la \lessactive(\var{A}, \var{P}, \var{Q}), \nota \paraminf(\var{A}, \var{P}, \var{Q}).
\end{align*}
Indeed, this constraint removes all parametrization results where parameters $K_{\var{A},\omega_\var{P}}$ and $K_{\var{A},\omega_\var{Q}}$ exist
such that $\var{A}$ is less activated by the set of resources $\omega_\var{P}$ than it is by $\omega_\var{Q}$,
but $K_{\var{A},\omega_\var{Q}} \ltsegm K_{\var{A},\omega_\var{P}}$,
thus violating the monotonicity assumption.
Of course, other assumptions can be formulated in the same way.

@@

Then, due to the fact that the parameters $K_{a,\{b\}} = \segm{1}{1}$ and $K_{a,\{c\}} = \segm{1}{1}$ could be inferred,
the monotonicity assumption (\pref{pro:param_enum_monotonicity}) removes all parametrizations with: $0 \in K_{a,\{a,b\}} \vee 0 \in K_{a,\{a,c\}}$.
In the end, the remaining possible values for the two parameters that were not inferred are:
\begin{align*}
  (K_{a,\{a,b\}} ; K_{a,\{a,c\}}) \in \{ \segm{1}{1} , \segm{2}{2} , \segm{1}{2} \}^2
\end{align*}
All these candidates respect the other properties of \pref{ssec:admissible-K} and no more candidates are filtered.
We thus obtain the results of \pref{ex:enum-param-runningPH-1}.
\end{example}

>>>>

\begin{example}\label{ex:asp-atom}
Consider a PH model such as \pref{ex:infer-param-runningPH-1}:
in order to state the existence of each component, we use a predicate called $\component$ with two arguments:
\begin{align*}
  &\component(x, n).
\end{align*}
where $x$ is the name of the component and $l_x = n$ is its maximum expression level.
\end{example}

An ASP program is a set of rules as described above.
Solving an ASP program means finding an \emph{answer set}, which is a minimal set of atoms that respect all the rules.
In order to formally define this notion of answer set,
let us define a \emph{definite rule} as a rule with no negation of atoms (noted “$\nota$” above),
and a \emph{definite program} as a program containing only definite rules.
Let $S$ be a set of atoms: a definite rule is \emph{satisfied} by $S$ if $\atom{H} \in S$,
or if $\exists i \in \segm{1}{n}, A_i \notin S$.
Given this definition of satisfaction, we define the answer set of a definite program $\Pi$
as the (unique) minimal set $S$ of atoms that satisfies all the rules in $\Pi$.

In order to consider the general case,
for all non-definite program $\Pi$ and set $S$ of atoms, we denote $\Pi^S$ the reduct of $\Pi$ w.r.t.~$S$, defined from $\Pi$ by:
\begin{enumerate}
  \item deleting all the rules that have a negation of atom $\nota \atom{B}_i$ in the body where $\atom{B}_i \in S$, and
  \item removing all negations of atoms in the bodies of the remaining rules.
\end{enumerate}
Then, a set of atoms $S$ is an answer set of a non-definite program $\Pi$ if it is the answer set of $\Pi^S$, which is a definite program.
We note that several answer sets can be solution to the same non-definite program, and in practice a solver can be directed to enumerate them all.

@@

Note that it is possible to define a simple rule with no $body$ part.
Such a rule is called a \emph{fact}, and its $head$ atom consequently has to belong to all answer sets.
For instance, the information describing the studied model (the original PH model and the inferred IG and parameters) are expressed in ASP using facts.

@@

\subsection{Variables}
To describe the sets of all expression levels of each component (\ie the set $\segm{0}{l_a}$ for each $a \in \Gamma$),
one can use atoms of the form $\componentlevels(a, k)$ to state that $k \in \segm{0}{l_a}$.
Variables here come in handy to enumerate each possible constant $k$ for each component $a$:
before solving, any rule containing variables is \emph{grounded}, that is, replaced by an equivalent set of rules with constants only.
%during the solving, any rule containing variables is duplicated in order to replace each variable by all the possible constants it could represent.
The following rule, for example, contains three variables ($\var{A}$, $\var{K}$ and $\var{M}$) and enumerates the set of possible expression levels of each component in the system:
\begin{align*}
  \componentlevels(A, K) \la \component(\var{A}, \var{M}), 0 \leq K \leq M.
\end{align*}
where the notation “$\leq$” stands for a shortcut in ASP which has the same meaning as the mathematical operator.

@@

\subsection{Constraints}\label{sssec:constraints}
Finally, a \emph{constraint} is a rule with no $head$ part:
\begin{align*}
  \la \atom{A}_1, \atom{A}_2, \dots, \atom{A}_n, \nota \atom{B}_1, \nota \atom{B}_2, \dots, \nota \atom{B}_m.
\end{align*}
A constraint is satisfied only if its $body$ is not satisfied,
which thus allows to invalidate answer sets containing some unwanted combinations of atoms.
In the scope of parameters enumeration, for example, constraints are especially useful to filter parametrizations
that contain non-interval parameters, or that do not respect the assumptions of \pref{ssec:admissible-K}.
Indeed, suppose that:
\begin{itemize}
  \item $\lessactive(a, p, q)$ states that $\omega_p$ is a set of resources of $a$ with (loosely) less activators and more inhibitors than $\omega_q$;
  \item $\paraminf(a, p, q)$ states that: $K_{a,\omega_p} \leqsegm K_{a,\omega_q}$.
\end{itemize}
Then, the monotonicity assumption (\pref{pro:param_enum_monotonicity}) is formulated as the following constraint:
\begin{align*}
  \la \lessactive(\var{A}, \var{P}, \var{Q}), \nota \paraminf(\var{A}, \var{P}, \var{Q}).
\end{align*}
Indeed, this constraint removes all parametrization results where parameters $K_{\var{A},\omega_\var{P}}$ and $K_{\var{A},\omega_\var{Q}}$ exist
such that $\var{A}$ is less activated by the set of resources $\omega_\var{P}$ than it is by $\omega_\var{Q}$,
but $K_{\var{A},\omega_\var{Q}} \ltsegm K_{\var{A},\omega_\var{P}}$,
thus violating the monotonicity assumption.
Of course, other assumptions can be formulated in the same way.

@@

Then, due to the fact that the parameters $K_{a,\{b\}} = \segm{1}{1}$ and $K_{a,\{c\}} = \segm{1}{1}$ could be inferred,
the monotonicity assumption (\pref{pro:param_enum_monotonicity} and expressed in ASP above) removes all parametrizations with:
$0 \in K_{a,\{a,b\}} \vee 0 \in K_{a,\{a,c\}}$.
In the end, the remaining possible values for the two parameters that were not inferred are:
\begin{align*}
  (K_{a,\{a,b\}} ; K_{a,\{a,c\}}) \in \{ \segm{1}{1} , \segm{2}{2} , \segm{1}{2} \}^2
\end{align*}
All these candidates respect the other properties of \pref{ssec:admissible-K} and no more candidates are filtered.
We thus obtain the results of \pref{ex:enum-param-runningPH-1}.
\end{example}

### 2013-08-03 parts-discussion.tex
all source files in the same folder (TCS submission)

We first propose an original inference of the Interaction Graph (IG) from a BRN
having its dynamics specified in the PH framework.
An IG gives a compact abstract representation of the influence of the components between each
others.
Then, based on a prior inference of Ren\'e Thomas' parametrization for BRNs from a PH model, we
delimit the set of compatible Thomas' parametrizations that are compatible with the PH dynamics,
and give arguments for their correctness.
A parametrization is compatible with the PH if its dynamics (in terms of possible transitions) is included in the PH dynamics.
The enumeration of such parametrizations is efficiently tackled using Answer Set Programming.
We illustrated the overall method with several results on large biological models.

Several extensions of the presented work are now to be considered.
First, the link between successively refined models of a system could be formally studied.
Indeed, it is convenient to refine a PH model by removing actions or adding cooperations;
the study of such process would allow to predict behavioral changes and lead to more accurate models.
Such work would also require further study on the inference of the IG in order to better understand and possibly avoid
the presence of some of the current spurious auto-edges.
Second, the inference of BRN multiplexes \cite{BernotMultiplexes} may be of practical interest 
as they allow to implicitly reduce the possible parametrizations by making cooperations appear
in the IG.
Because of its atomicity, the PH allows to specify a range of cooperations that cannot be
completely captured by a single instance of BRN multiplexes, then encouraging the inference of a set
of compatible ones.
Finally, in order to improve the performances in the IG inference, we will consider projection operations on
the PH structure to undo cooperations between components and reduce the cardinality of
configurations to explore by making the interactions independent.

>>>>

We first propose an original inference of the Interaction Graph (IG) from a BRN
having its dynamics specified in the PH framework.
An IG gives a compact abstract representation of the influence of the components between each
others.
Then, based on a prior inference of Ren\'e Thomas' parametrization for BRNs from a PH model, we
delimit the set of admissible Thomas' parametrizations that are compatible with the PH dynamics,
and give arguments for their correctness.
A parametrization is compatible with the PH if its dynamics (in terms of possible transitions) is included in the PH dynamics.
The enumeration of such parametrizations is efficiently tackled using Answer Set Programming.
We illustrate the overall method with several results on large biological models.

Several extensions of the presented work are now to be considered.
First, the link between successively refined models of a system could be formally studied.
Indeed, it is convenient to refine a PH model by removing actions or adding cooperations;
the study and formalization of such process would allow to predict behavioral changes and lead to more accurate models.
Such work would also require further study on the inference of the IG in order to better understand and possibly avoid
the presence of some of the current spurious auto-edges.
Second, the inference of BRN multiplexes \cite{BernotMultiplexes} may be of practical interest 
as they allow to implicitly reduce the possible parametrizations by making cooperations appear
in the IG.
Because of its atomicity, the PH allows to specify a range of cooperations that cannot be
completely captured by a single instance of BRN multiplexes, then encouraging the inference of a set
of compatible ones.
Finally, in order to improve the performances in the IG inference, we will consider projection operations on
the PH structure to undo cooperations between components and reduce the cardinality of
configurations to explore by making the interactions independent.

### 2013-08-03 parts-grn.tex
all source files in the same folder (TCS submission)

\paragraph{Interaction Graph of the dynamics}
From the dynamics specification of a BRN, one can infer back the IG that contain only the functional
regulations.
Let us define
$f^a(s) = s[a]$ if $s[a]\in K_{a, \GRNres{a}{s}}$, 
$f^a(s) = s[a]+1$ if $s[a] < K_{a, \GRNres{a}{s}}$, 
and 
$f^a(s) = s[a]-1$ if $s[a] > K_{a, \GRNres{a}{s}}$.
A positive (resp. negative) interaction from $b$ to $a$ is inferred if there exists a state $s$ such
that increasing $b$ in $s$ would increase (resp. decrease) $f^a$:
$f^a(s\{b_i\}) < \text{ (resp. $>$) } f^a(s\{b_{i+1}\})$, 
where $i < l_b$ and $s\{b_i\}$ denotes the state $s$ where the component $b$ is assigned to $b_i$.
Then, such an IG can be used to derive global properties on the dynamics
(e.g., \cite{Richard2010378,PR11-SASB}).

\paragraph{Discussion on maximum level restrictions}
The maximum level $l_a$ of a component $a$ is usually chosen equal to the number of outgoing edges of $a$ in the IG, \ie the number of components that $a$ regulates.
Indeed, each discrete expression level in $\segm{0}{l_a}$ stands for an arbitrary set of concentration levels for which $a$ has a certain influence on several other components it regulates.
Allowing more expression levels makes several levels useless as they represent the same influence of $a$ than some other levels;
on the contrary, reducing the maximum level requires that several thresholds on outgoing edges of $a$ share the same value, which is biologically unlikely.
However, these considerations are usually taken into account when one has to directly build a model in this framework, but are not mandatory.
In the scope of this paper, we allow any number of levels for components, without considering the number of outgoing edges,
as the number of processes in a PH sort is not constrained in any way.
For instance, in the example of \pref{fig:runningBRN},
it would be more relevant to have $l_a = 1$ and $\GRNedgef{a}{-}{1}{b} \in E$, as $a$ has only one outgoing edge,
but the value $l_a = 2$ makes the example more interesting, therefore $\GRNedgef{a}{-}{2}{b} \in E$ can be arbitrarily chosen.

>>>>

\paragraph{Interaction Graph of the dynamics}
From the dynamics specification of a BRN, one can infer back the IG that contain only the functional
regulations.
Let us define
$f^a(s) = s[a]$ if $s[a]\in K_{a, \GRNres{a}{s}}$
and
$f^a(s) = s[a]+\delta^a(s)$ otherwise.
A positive (resp. negative) interaction from $b$ to $a$ is inferred if there exists a state $s$ such
that increasing $b$ in $s$ would increase (resp. decrease) $f^a$; in other words, if
there exists a state $s$ so that:
$f^a(s\{b_i\}) < \text{(resp. $>$) } f^a(s\{b_{i+1}\})$, 
where $i < l_b$ and $s\{b_i\}$ denotes the state $s$ where the component $b$ is assigned to $b_i$.
Then, such an IG can be used to derive global properties on the dynamics
(e.g., \cite{Richard2010378,PR11-SASB}).

\paragraph{Discussion on maximum level restrictions}
The maximum level $l_a$ of a component $a$ is usually chosen equal to the number of outgoing edges of $a$ in the IG, \ie the number of components that $a$ regulates.
Indeed, each discrete expression level in $\segm{0}{l_a}$ stands for an arbitrary set of concentration levels of the protein coded by $a$
and for which it has a certain influence on several other components that it regulates.
Allowing more expression levels makes several levels useless as they represent the same influence of $a$ than some other levels;
on the contrary, reducing the maximum level requires that several thresholds on outgoing edges of $a$ share the same value, which is biologically unlikely.
However, these considerations are usually taken into account when one has to directly build a model in this framework, but are not mandatory.
In the scope of this paper, we allow any number of levels for components, without considering the number of outgoing edges,
as the number of processes in a PH sort is not constrained in any way.
For instance, in the example of \pref{fig:runningBRN},
it would be more relevant to have $l_a = 1$ and $\GRNedgef{a}{-}{1}{b} \in E$, as $a$ has only one outgoing edge,
but the value $l_a = 2$ makes the example more interesting; this is why $\GRNedgef{a}{-}{2}{b} \in E$ can be arbitrarily chosen.

### 2013-08-03 parts-infer-ig.tex
all source files in the same folder (TCS submission)

Finally, \pref{pro:wf-ph} sums up the conditions for a Process Hitting to be suitable for IG
inference.
In addition of having either component sorts or well-formed cooperative sorts, we also require that
there is no cycle between cooperative sorts, and that
sorts being never hit (\ie{} serving as an invariant environment) are components.

@@

At this point we can divide the set of sorts $\PHs$ into components ($\Gamma$, see \pref{eq:PH-components}) and cooperative sorts
($\PHs \setminus \Gamma$) that will not appear in the IG. 
We define in \pref{eq:ph_predec} the set of predecessors of a sort $a$, that is, the sorts influencing $a$
by considering direct actions and possible intermediate cooperative sorts.
The predecessors of $a$ that are components are the regulators of $a$, denoted $\PHpredecgene{a}$
(\pref{eq:regulators}).
\begin{align}
\begin{split}
\forall a \in \PHs, \PHpredec{a} &\DEF \{b \in \PHs \mid \exists n \in \mathbb{N}^*, \exists
(c^k)_{k \in \segm{0}{n}} \in \PHs^{n+1}, \\
                                   & \quad \quad c^0 = b \wedge c^n = a \\
                                   & \quad \quad \wedge \forall k \in \segm{0}{n-1},
                   c^k \in \PHdirectpredec{c^{k+1}} \cap (\PHs\setminus\Gamma)\}
\end{split}
\label{eq:ph_predec}
\\
\forall a\in \PHs, \PHpredecgene{a} & \DEF \PHpredec{a} \cap \Gamma
\label{eq:regulators}
\end{align}

Given a set $g$ of components and a configuration (\ie a sub-state) $\sigma$, $\ctx_g(\sigma)$
refers to the set of processes hitting $a$ regulated by any sort in $g$ (\pref{eq:ctx-sigma}).
If $g=\{b\}$, we simple note $\ctx_b(\sigma)$.
This set is composed of the active processes of sorts in $g$, and the focal process (assumed
unique) of the cooperative sorts $\upsilon$ hitting $a$ that have a predecessor in $g$.
The evaluation of the focal process of $\upsilon$ in context $\sigma$, denoted $\upsilon(\sigma)$,
relies on \pref{pro:wf-cooperative-sort}, which gives its value when all the direct predecessors of
$\upsilon$ are defined in $\sigma$.
When a predecessor $\upsilon'$ is not in $\sigma$, we extend the evaluation by recursively computing
the focal value of $\upsilon'$ is $\sigma$, as stated in \pref{eq:cooperative-eval},
where $\uplus$ stands for a union of states (which are sets of processes).
Because there is no cycle between cooperative sorts, this recursive evaluation of $\upsilon(\sigma)$
always terminates.
\begin{align}
\forall g\subset \Gamma,
  \ctx_g(\sigma) & \DEF \{ \sigma[b] \mid b\in g \} \cup \{ \upsilon(\sigma) \mid
\upsilon\in\PHdirectpredec{a} \setminus \Gamma \wedge g\cap \PHpredecgene{\upsilon} \neq \emptyset \}
\label{eq:ctx-sigma}
\\
\upsilon(\sigma) & \DEF
\upsilon(\sigma \uplus \state{\upsilon'(\sigma) \mid 
  \upsilon'\in\PHdirectpredec{\upsilon} \wedge
  \upsilon'\in\PHs\setminus\Gamma })
\label{eq:cooperative-eval}
\end{align}

@@

\pref{pps:inference-edges} details the inference of all existing influences between components occurring
with a threshold $t$.
The main idea of this inference is that
the presence of a positive (negative) influence of a component $b$ on $a$ denotes the fact that
there exists a state in which increasing the level of $b$ tends to make the future level of $a$ rise (drop)
(\pref{eq:edges-inference}).
Therefore, these influences are split into positive and negative ones, and represent possible edges in the final IG.
Furthermore, studying the influences of the groups of regulators of $a$
allows to study its auto-influences, and thus infer auto-edges on $a$ in the IG (\pref{eq:edges-inference-auto}).
Finally, \pref{eq:edges-inference-noreg} handles the special case where $a$ has no regulators.
We ignore the cases where a component has no visible influence on another.
%
\begin{proposition}[Influences inference]\label{pps:inference-edges}
We define the set of positive (resp. negative) influences $\hat{E}_+$ (resp. $\hat{E}_-$) for any $a\in\Gamma$ by:
\begin{align}
% Arcs a -> b, a ≠ b
\begin{split}\label{eq:edges-inference}
  \forall b\in\PHpredecgene{a}, \forall s \in \{ +, - \}, \\
  b \xrightarrow{t+1} a \in \hat{E}_s \Longleftrightarrow\ & \exists g \in X(a), b \in g,
  \exists \sigma \in \textstyle\prod_{c\in g \cup \{ a \}} \PHl_c, \\
    &\qquad \{ b_t, b_{t+1} \} \subset \PHl_b \wedge b_t \in \sigma,\\
    &\qquad \exists a_j \in \irB_a(\sigma), \exists a_k \in \irB_a(\sigma\{b_{t+1}\}), \\
    &\qquad s = \f{sign}(k - a)
\end{split}\\
% Auto-arcs
\begin{split}\label{eq:edges-inference-auto}
  \forall s \in \{ +, - \}, \quad\qquad\qquad \\
  a \xrightarrow{t+1} a \in \hat{E}_s \Longleftrightarrow\ & \exists g \in X(a), \exists b \in g,
  \exists \sigma \in \textstyle\prod_{c\in g \cup \{ a \}} \PHl_c, \\
    &\qquad \{ b_t, b_{t+1} \} \subset \PHl_b \wedge b_t \in \sigma,\\
    &\qquad \exists a_j \in \irB_a(\sigma), \exists a_k \in \irB_a(\sigma\{b_{t+1}\}), \\
    &\qquad s = \f{sign}(a_k - a_j)
\end{split}\\
\begin{split}\label{eq:edges-inference-noreg}
  \forall s \in \{ +, - \}, \quad\qquad\qquad \\
  a \xrightarrow{t+1} a \in \hat{E}_s \Longleftrightarrow\ & \reg{a} = \emptyset \wedge \{ a_t, a_{t+1} \} \subset \PHl_a, \\
    &\qquad \exists a_j \in \irB_a(\PHetat{a_t}), \exists a_k \in \irB_a(\PHetat{a_{t+1}}), \\
    &\qquad s = \f{sign}(a_k - a_j)
\end{split}
\end{align}
where $\f{sign}(x) = \begin{cases}|x|/x & \text{ if $x \neq 0$} \\ 0 & \text{ if $x = 0$}\end{cases}$.
\end{proposition}

>>>>

Finally, \pref{pro:wf-ph} sums up the conditions for a Process Hitting to be suitable for IG
inference.
In addition of having either component sorts or well-formed cooperative sorts, we also require that
there is no cycle between cooperative sorts, and that
sorts being never hit (\ie serving as an invariant environment) are components.

@@

At this point we can divide the set of sorts $\PHs$ into components ($\Gamma$, see \pref{eq:PH-components}) and cooperative sorts
that will not appear in the IG ($\PHs \setminus \Gamma$).
We define in \pref{eq:ph_predec} the set of predecessors of a sort $a$, that is, the sorts influencing $a$
by considering direct actions and possible intermediate cooperative sorts.
The predecessors of $a$ that are components are the regulators of $a$, denoted $\PHpredecgene{a}$
(\pref{eq:regulators}).
\begin{align}
\begin{split}
\forall a \in \PHs, \PHpredec{a} &\DEF \{b \in \PHs \mid \exists n \in \mathbb{N}^*, \exists
(c^k)_{k \in \segm{0}{n}} \in \PHs^{n+1}, \\
                                   & \quad \quad c^0 = b \wedge c^n = a \\
                                   & \quad \quad \wedge \forall k \in \segm{0}{n-1},
                   c^k \in \PHdirectpredec{c^{k+1}} \cap (\PHs\setminus\Gamma)\}
\end{split}
\label{eq:ph_predec}
\\
\forall a\in \PHs, \PHpredecgene{a} & \DEF \PHpredec{a} \cap \Gamma
\label{eq:regulators}
\end{align}

Given a set $g$ of components and a configuration (\ie a sub-state) $\sigma$, $\ctx_g(\sigma)$
refers to the set of processes hitting $a$ regulated by any sort in $g$ (\pref{eq:ctx-sigma}).
If $g=\{b\}$, we simply note $\ctx_b(\sigma)$.
This set is composed of the active processes of sorts in $g$, and the focal process (assumed
unique) of the cooperative sorts $\upsilon$ hitting $a$ that have a predecessor in $g$.
The evaluation of the focal process of $\upsilon$ in context $\sigma$, denoted $\upsilon(\sigma)$,
relies on \pref{pro:wf-cooperative-sort}, which gives its value when all the direct predecessors of
$\upsilon$ are defined in $\sigma$.
When a predecessor $\upsilon'$ is not in $\sigma$, we extend the evaluation by recursively computing
the focal value of $\upsilon'$ is $\sigma$, as stated in \pref{eq:cooperative-eval},
where $\uplus$ stands for a union of states (which are sets of processes).
Because there is no cycle between cooperative sorts, this recursive evaluation of $\upsilon(\sigma)$
always terminates.
\begin{align}
\forall g\subset \Gamma,
  \ctx_g(\sigma) & \DEF \{ \sigma[b] \mid b\in g \} \cup \{ \upsilon(\sigma) \mid
\upsilon\in\PHdirectpredec{a} \setminus \Gamma \wedge g\cap \PHpredecgene{\upsilon} \neq \emptyset \}
\label{eq:ctx-sigma}
\\
\upsilon(\sigma) & \DEF
\upsilon(\sigma \uplus \state{\upsilon'(\sigma) \mid 
  \upsilon'\in\PHdirectpredec{\upsilon} \wedge
  \upsilon'\in\PHs\setminus\Gamma })
\label{eq:cooperative-eval}
\end{align}

@@

\pref{pps:inference-edges} details the inference of all existing influences between components occurring
with a threshold $t$.
The main idea of this inference is that
the presence of a positive (negative) influence of a component $b$ on $a$ denotes the fact that
there exists a state in which increasing the level of $b$ tends to make the future level of $a$ rise (drop)
(\pref{eq:edges-inference}).
Therefore, these influences are split into positive and negative ones, and represent possible edges in the final IG.
Furthermore, studying the influences of the groups of regulators of $a$
allows to study its auto-influences, and thus infer auto-edges on $a$ in the IG (\pref{eq:edges-inference-auto}).
Finally, \pref{eq:edges-inference-noreg} handles the special case where $a$ has no regulators.
We ignore the cases where a component has no visible influence on another.
%
\begin{proposition}[Influences inference]\label{pps:inference-edges}
We define the set of positive (resp. negative) influences $\hat{E}_+$ (resp. $\hat{E}_-$) for any $a\in\Gamma$ by:
% Arcs a -> b, a ≠ b
\begin{align}
\begin{split}\label{eq:edges-inference}
  \forall b\in\PHpredecgene{a}, \forall s \in \{ +, - \}, \\
  b \xrightarrow{t+1} a \in \hat{E}_s \Longleftrightarrow\ & \exists g \in X(a), b \in g,
  \exists \sigma \in \textstyle\prod_{c\in g \cup \{ a \}} \PHl_c, \\
    &\qquad \{ b_t, b_{t+1} \} \subset \PHl_b \wedge b_t \in \sigma,\\
    &\qquad \exists a_j \in \irB_a(\sigma), \exists a_k \in \irB_a(\sigma\{b_{t+1}\}), \\
    &\qquad s = \f{sign}(k - a)
\end{split}
\end{align}
% Auto-arcs depuis les groupes de régulateurs
\begin{align}
\begin{split}\label{eq:edges-inference-auto}
  \forall s \in \{ +, - \}, \quad\qquad\qquad \\
  a \xrightarrow{t+1} a \in \hat{E}_s \Longleftrightarrow\ & \exists g \in X(a), \exists b \in g,
  \exists \sigma \in \textstyle\prod_{c\in g \cup \{ a \}} \PHl_c, \\
    &\qquad \{ b_t, b_{t+1} \} \subset \PHl_b \wedge b_t \in \sigma,\\
    &\qquad \exists a_j \in \irB_a(\sigma), \exists a_k \in \irB_a(\sigma\{b_{t+1}\}), \\
    &\qquad s = \f{sign}(a_k - a_j)
\end{split}
\end{align}
% Auto-arcs des composants sans prédécesseurs
\begin{align}
\begin{split}\label{eq:edges-inference-noreg}
  \forall s \in \{ +, - \}, \quad\qquad\qquad \\
  a \xrightarrow{t+1} a \in \hat{E}_s \Longleftrightarrow\ & \reg{a} = \emptyset \wedge \{ a_t, a_{t+1} \} \subset \PHl_a, \\
    &\qquad \exists a_j \in \irB_a(\PHetat{a_t}), \exists a_k \in \irB_a(\PHetat{a_{t+1}}), \\
    &\qquad s = \f{sign}(a_k - a_j)
\end{split}
\end{align}
where $\f{sign}(x) = \begin{cases}|x|/x & \text{ if $x \neq 0$} \\ 0 & \text{ if $x = 0$}\end{cases}$ \enspace.
\end{proposition}

### 2013-08-03 parts-infer-param.tex
all source files in the same folder (TCS submission)

\begin{example}
Regarding the refined PH of \pref{fig:runningPH-2}, all parameters can be inferred.
We obtain the same value for the inferred parameters of \pref{ex:infer-param-runningPH-1},
together with the following results:
\begin{align*}
  K_{a,\{a,b\}} &= 1 &
  K_{a,\{a,c\}} &= 1
\end{align*}
\end{example}

Given the \pref{pps:param_K}, we see that in some cases, the inference of the targeted parameter is impossible.
This can be due to a lack of cooperation between regulators: when two regulators independently hit a component, their actions can have opposite effects, leading to either an indeterministic evolution or to oscillations.
Such an indeterminism is not possible in a BRN as in a given configuration of regulators, a component can only have an interval attractor, and eventually reaches a steady-state.
In order to avoid such inconclusive cases, one has to ensure that no such behavior is allowed by
either removing undesired actions or using cooperative sorts to prevent opposite influences between
regulators.

@@

When building a BRN, one has to find the parametrization that best describes the desired behavior of the studied system.
Complexity is inherent to this process as the number of possible parametrizations for a given IG is exponential w.r.t. the number of components.
However, the method of parameters inference presented in this section gives some information about necessary parameters given a certain dynamics described by a PH.
This information thus drops the number of possible parametrizations, allowing to find the desired behavior more easily.

>>>>

\begin{example}
Regarding the refined PH of \pref{fig:runningPH-2}, all parameters can be inferred.
We obtain the same value for the inferred parameters of \pref{ex:infer-param-runningPH-1},
together with the following results:
\begin{align*}
  K_{a,\{a,b\}} &= \segm{1}{1} &
  K_{a,\{a,c\}} &= \segm{1}{1}
\end{align*}
\end{example}

Given the \pref{pps:param_K}, we see that in some cases, the inference of the targeted parameter is impossible.
This can be due to a lack of cooperation between regulators:
when two regulators independently hit a component, their actions can have opposite effects, leading to two possible evolutions.
Such an indeterminism is not possible in Thomas modeling as in a given configuration of regulators,
a component can only have an interval attractor, and can therefore evolve in only one direction.
In order to avoid such inconclusive cases, one has to ensure that no such behavior is allowed by
either removing undesired actions or using cooperative sorts to prevent opposite influences between
concurrent regulators.

@@

When building a BRN, one has to find the parametrization that best describes the desired behavior of the studied system.
Complexity is inherent to this process as the number of possible parametrizations for a given IG is exponential w.r.t.~the number of components.
However, the method of parameters inference presented in this section gives some information about necessary parameters given a certain dynamics described by a PH.
This information thus drops the number of possible parametrizations, allowing to find the desired behavior more easily.
